Prosper Loan Exploratory Data Analysis by Brandon Brown
========================================================

> Prosper is a peer-to-peer lending marketplace in the United States which 
provides a platform for borrowers and lenders (Reference 1). Borrowers can 
apply for a fixed-rate, fixed-term loan of between $2,000 and $35,000 and individuals/institutions can invest in loans which generate attractive returns (Reference 1). Data is available on Prosper loans (Reference 2). This data 
includes interest rates, borrower employment status, borrower income, loan 
amount, borrower credit history and many other variables (Reference 3). The 
purpose of this project is to explore the data to determine which conditions 
might be associated with whether a borrower will be successful in paying back 
the original loan amount and interest. I expect to see which borrower employment status, credit score, debt to income ratio and other loan variables are most associated with those who default on their loans and who successfully complete 
their loan payments.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading all of the packages I will use in this project. I will be using 
# ggplot2 to construct plots (Reference 1), knitr to put the R markdown document into an HTML document (Reference 2) and dplyr to mutate and transform the data (Reference 3).  I had to add a repos section to get the knitting of the html 
# file to complete (Reference 4).  

# NOTE ABOUT REFERENCES: At the end of this RMD file I have two Resources 
# sections.  The first is inside of an R chunck and is only for code inside R chuncks.  The second is outside the R chunck and is only for analyses and 
# paragraphs outside the R chuncks.  I did this so that references for code 
# inside the R chuncks wouldn't be visible in a knitted HTML file if users 
# want to knit together the file (because the R chuncks themselves won't be visible). 


install.packages("ggplot2", dependencies = T, 
                 repos = "http://cran.us.r-project.org") 
install.packages("knitr", dependencies = T, 
                 repos = "http://cran.us.r-project.org")
install.packages("dplyr", dependencies = T, 
                 repos = "http://cran.us.r-project.org")

# I set echo, message and warning to all be false so that I wouldn't see the 
# code in the HTML document or any warning messages (Reference 5)

# Load the libraries using the library function.  Note that you can use ? in 
# front of any command I use to see what the command does.  This tells what the command does (Reference 6).

library(ggplot2)
library(knitr)
library(dplyr)

# Get and set working directory (Reference 7)
# NOTE: For users wishing to contribute, you will need to set the path to that 
# you have cloned the project to inside the parentheses in setwd.
getwd()

# I used the paste function because my file name was too long and I needed
# to shorten it to below 80 characters (Reference 8).
setwd(paste("C:/Users/brownb/Desktop/Udacity/Data Analyst Nanodegree/",
            "Term 2/Exploratory Data Analysis/Project 2", sep = ""))
```

```{r echo=FALSE, Load_the_Data}
# Load the Data (Reference 9)

prosper_df <- read.csv('prosperLoanData.csv')
```

# Univariate Analysis

First, I will perform some univariate analyses, plotting bar charts and 
histograms of various variables of interest to get an idea about how each 
variable is related to measures such as the number of borrowers who default and 
the number who successfully pay back their loans.  Once we have the univariate plots, we can have an idea about what variables we may want to explore in a multi-variate fashion.

```{r echo=FALSE, Univariate_Plots}
# First, I need to see the dimensions and structure of the data as is. This way
# I can get an idea at any transformations I need to perform on the data.

# Obtain the dimensions of the dataframe as is (Reference 10).
dim(prosper_df)

# Obtain the structure of the dataframe as is (Reference 11).
str(prosper_df)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, resources}
# Resources - code

# Reference 1: http://ggplot2.org/

# Reference 2: https://yihui.name/knitr/

# Reference 3: https://dplyr.tidyverse.org/

# Reference 4: https://stackoverflow.com/questions/29189663/trying-to-
# publish-an-r-notebook-and-keep-getting-the-same-error-error-in-contri

# Reference 5: http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html

# Reference 6: https://stackoverflow.com/questions/5595512/what-is-the-
# difference-between-require-and-library

# Reference 7: http://rfunction.com/archives/1001

# Reference 8: https://stackoverflow.com/questions/6329962/split-code-over-
# multiple-lines-in-an-r-script

# Reference 9: http://rprogramming.net/read-csv-in-r/

# Reference 10: https://stat.ethz.ch/R-manual/R-devel/library/base/html/dim.html

# Reference 11: https://www.rdocumentation.org/packages/utils/versions/3.4.3/
# topics/str

# NOTE: This is the Resources list only for the code chuncks.  I am placing this list inside of this chunck so that they do not appear when knitting the HTML 
# file (for users who want to do that at the end). The only resources I want to appear when knitting the HTML file are those referenced inside the file and 
# meant to be visible (outside the R chuncks).
```

# Resources

Reference 1: https://www.prosper.com/home/about

Reference 2: https://www.kaggle.com/jschnessl/prosperloans/feed

Reference 3: https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0