Prosper Loan Exploratory Data Analysis by Brandon Brown
========================================================

> Prosper is a peer-to-peer lending marketplace in the United States which provides a platform for borrowers and lenders (Reference 1). Borrowers can apply for a fixed-rate, fixed-term loan of between $2,000 and $35,000 and individuals/institutions can invest in loans which generate attractive returns (Reference 1). Data is available on Prosper loans (Reference 2). This data includes interest rates, borrower employment status, borrower income, loan amount, borrower credit history and many other variables (Reference 3). The purpose of this project is to explore the data to determine which conditions might be associated with whether a borrower will be successful in paying back the original loan amount and interest. I expect to see which borrower incomes, employment status, credit history and other variables are most associated with those who default on their loans and who successfully complete their loan payments.



```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading all of the packages I will use in this project. I will be using ggplot2 to construct plots (Reference 4), knitr to put the R markdown document into an HTML document (Reference 5) and dplyr to mutate and transform the data (Reference 6).  I had to add a repos section to get the knitting of the html file to complete (Reference 7).  

install.packages("ggplot2", dependencies = T, repos = "http://cran.us.r-project.org") 
install.packages("knitr", dependencies = T, repos = "http://cran.us.r-project.org")
install.packages("dplyr", dependencies = T, repos = "http://cran.us.r-project.org")

# I set echo, message and warning to all be false so that I wouldn't see the code in the HTML document or any warning messages (Reference 8)

# Load the libraries using the library function.  Note that you can use ? in front of any command I use to see what the command does.  This tells what the command does (Reference 9).

library(ggplot2)
library(knitr)
library(dplyr)

# Get and set working directory (Reference 10)
# NOTE: For users wishing to contribute, I have left what is inside the parentheses blank.  You will need to add the path that you have cloned the project to inside the parentheses in setwd.
getwd()
setwd("")
```

```{r echo=FALSE, Load_the_Data}
# Load the Data (Reference 11)

prosper_df <- read.csv('prosperLoanData.csv')
```

# Univariate Plots

> First, I will perform some univariate analyses, plotting bar charts and histograms of various variables of interest to get an idea about how each variable is related to measures such as the number of borrowers who default and the number who successfuly pay back their loans.  Once we have the univariate plots, we can have an idea about what variables we may want to explore in a multi-variate fashion.




# Resources

> Reference 1: https://www.prosper.com/home/about

> Reference 2: https://www.kaggle.com/jschnessl/prosperloans/feed

> Reference 3: https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0

> Reference 4: http://ggplot2.org/

> Reference 5: https://yihui.name/knitr/

> Reference 6: https://dplyr.tidyverse.org/

> Reference 7: https://stackoverflow.com/questions/29189663/trying-to-publish-an-r-notebook-and-keep-getting-the-same-error-error-in-contri

> Reference 8: http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html

> Reference 9: https://stackoverflow.com/questions/5595512/what-is-the-difference-between-require-and-library

> Reference 10: http://rfunction.com/archives/1001

> Reference 11: http://rprogramming.net/read-csv-in-r/