Prosper Loan Exploratory Data Analysis by Brandon Brown
========================================================

> Prosper is a peer-to-peer lending marketplace in the United States which 
provides a platform for borrowers and lenders (Reference 1). Borrowers can 
apply for a fixed-rate, fixed-term loan of between $2,000 and $35,000 and individuals/institutions can invest in loans which generate attractive returns (Reference 1). Data is available on Prosper loans (Reference 2). This data 
includes interest rates, borrower employment status, borrower income, loan 
amount, borrower credit history and many other variables (Reference 3). The 
purpose of this project is to explore the data to determine which conditions 
might be associated with whether a borrower will be successful in paying back 
the original loan amount and interest. I expect to see which borrower employment status, credit score, debt to income ratio and other loan variables are most associated with those who default on their loans and who successfully complete 
their loan payments.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading all of the packages I will use in this project. I will be using 
# ggplot2 to construct plots (Reference 1), knitr to put the R markdown document into an HTML document (Reference 2) and dplyr to mutate and transform the data (Reference 3).  I had to add a repos section to get the knitting of the html 
# file to complete (Reference 4).  

# NOTE ABOUT REFERENCES: At the end of this RMD file I have two Resources 
# sections.  The first is inside of an R chunk and is only for code inside R chunks.  The second is outside the R chunk and is only for analyses and 
# paragraphs outside the R chunks.  I did this so that references for code 
# inside the R chunks wouldn't be visible in a knitted HTML file if users 
# want to knit together the file (because the R chunks themselves won't be visible). 


install.packages("ggplot2", dependencies = T, 
                 repos = "http://cran.us.r-project.org") 
install.packages("knitr", dependencies = T, 
                 repos = "http://cran.us.r-project.org")
install.packages("dplyr", dependencies = T, 
                 repos = "http://cran.us.r-project.org")

# I set echo, message and warning to all be false so that I wouldn't see the 
# code in the HTML document or any warning messages (Reference 5)

# Load the libraries using the library function.  Note that you can use ? in 
# front of any command I use to see what the command does.  This tells what the command does (Reference 6).

library(ggplot2)
library(knitr)
library(dplyr)

# Get and set working directory (Reference 7)
# NOTE: For users wishing to contribute, you will need to set the path to that 
# you have cloned the project to inside the parentheses in setwd.
getwd()

# I used the paste function because my file name was too long and I needed
# to shorten it to below 80 characters (Reference 8).
setwd(paste("C:/Users/brownb/Desktop/Udacity/Data Analyst Nanodegree/",
            "Term 2/Exploratory Data Analysis/Project 2", sep = ""))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data (Reference 9)

prosper_df <- read.csv('prosperLoanData.csv')
```

# Univariate Analysis

First, I will perform some univariate analyses, plotting bar charts and 
histograms of various variables of interest to get an idea about how each 
variable is related to measures such as the number of borrowers who default and 
the number who successfully pay back their loans.  Once we have the univariate plots, we can have an idea about what variables we may want to explore in a multi-variate fashion.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
# First, I need to see the dimensions and structure of the data as is. This way
# I can get an idea at any transformations I need to perform on the data.

# Obtain the dimensions of the dataframe as is (Reference 10).
dim(prosper_df)

# Obtain the structure of the dataframe as is (Reference 11).
str(prosper_df)
```

  Since there are 81 variables, it is impossible to examine the relationships between all of them.  So I will pick a subset of the variables.  My main 
question is how variables of interest relate to the ability or inability of a borrower to pay back the loan amount plus interest.  Because of this, the most important variable I will be looking at is Loan Status.  This is a categorical variable with categories such as Completed, Defaulted, and Past Due for varying number of days among others.  It would make sense that a person's employment 
status will affect their ability to pay back the loans so I will also look at 
this variable.  A borrower's credit score is a statistical method to determine 
the likelihood that they will pay back a loan in a specified amount of time (Reference 4), so I will look at that variable and see how it is related to loan status (specifically the lower range of a person's credit score at the time the listing was created).  It would be expected that a borrower with a low debt to income ratio would have an easier time paying back loans, so this variable will 
be considered in this analysis.  
  
  Two variables which are only for listings post July 2009 are Prosper Score and Prosper Rating.  A Prosper Score is a number from 1 to 11 determined by Prosper which represents the probability of a loan going "bad" where bad is defined as 
the loan going 60 or more days past due within 12 months of the loan origination 
(1 is the highest risk score and 11 is the lowest risk) (Reference 5).  I will 
see how this variable is related to loan status and also to another variable, 
the number of days the loan is delinquent (number of days a payment is past due) (Reference 6).  A Prosper Rating is a grade assigned to a loan to determine the estimated annual loss to potential investors (AA to HR with AA having the lowest estimated loss and HR having the highest) (Reference 7).  It would be expected 
that in general, loans with the best rating (AA) would have the fewest number of borrowers who default on the loan.  One important point to keep in mind during 
this project is that a Prosper Rating may not be accurate because it is based on 
a borrower's credit report which may contain inaccurate information (Reference 
7). Another variable that will be examined is whether or not a borrower is a 
home owner.  If a borrower is a home owner, I would expect they will have higher credit worthiness and be less susceptible to going into default.  Finally, I 
will look at Borrower APR.  An APR is an annual charged rate which is the cost 
of borrowing the money (Reference 8).  My guess is that the lower this amount 
is, the easier it will be to pay back the loans.  I will look at this variable 
as well. 

  In summary the independent variables for this project are loan status and 
number of days the current loan is delinquent.  The dependent variables are employment status, credit score range lower, debt to income ratio, prosper 
score, prosper rating, borrower APR and is borrower a homeowner.  Note that for prosper score and prosper rating I will need to subset the data frame to only include loans which actually have a score and rating (those post July 2009). 

```{r echo=FALSE, message=FALSE, warning=FALSE, subset_dataframe}

# Subset the dataframe to only include the variables I described above

# Create a vector of strings representing the variable names from the prosper
# dataframe I wish to keep (Reference 12).
selected_variables <- c('LoanStatus', 'EmploymentStatus', 'BorrowerAPR',
                        'CreditScoreRangeLower', 'DebtToIncomeRatio',
                        'ProsperScore','ProsperRating..Alpha.',
                        'LoanCurrentDaysDelinquent', 'IsBorrowerHomeowner')


# Create new dataframe with only the variables I wish to keep (Reference 13).
prosp_sub <- subset(prosper_df, select = selected_variables)

# Rename the variables (Reference 14)

prosp_sub <- rename(prosp_sub, loan_status = LoanStatus, 
       employment_status = EmploymentStatus, borrower_apr = BorrowerAPR,
       credit_score_range_lower = CreditScoreRangeLower,
       debt_to_income_ratio = DebtToIncomeRatio, prosper_score = ProsperScore,
       prosper_rating = ProsperRating..Alpha., 
       days_delinquent = LoanCurrentDaysDelinquent, 
       is_homeowner = IsBorrowerHomeowner)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, loan_status_graph}

# Plotting a bar graph of loan_status; proportions for each category

# After plotting an initial version of the bar graph for loan_status, I 
# realized that there were several variables of Past Due.  I will now combine 
# all these variables into one variable (combine any Past Due amounts to simply 
# be past due). Since I mainly only care about statuses of Defaulted and
# Completed.

# Create a variable which contains all the possible Past Due categories.  
# They will be lumped into one category called "Past Due" (Reference 12)

past_due <- c('Past Due (>120 days)', 'Past Due (1-15 days)', 
              'Past Due (16-30 days)', 'Past Due (31-60 days)', 
              'Past Due (61-90 days)', 'Past Due (91-120 days)')

# First I need to add a factor level to the factor variable loan_status 
# (Reference 15).

prosp_sub$loan_status <- factor(prosp_sub$loan_status, 
                    levels = c(levels(prosp_sub$loan_status), "Past Due"))

# Now for all categories contained in past_due, set them to 'Past Due'
# (Reference 16)

prosp_sub[prosp_sub$loan_status %in% past_due, 1] <- c('Past Due')

# I'm now goint to lump "FinalPaymentInProgress" with Completed since loans
# in this status are basically complete.

# Create variable called complete with all categories that will be made
# to be "Complete" (Reference 12)
complete <- c('Completed', 'FinalPaymentInProgress')

# Now for all categories contained in complete, set them to 'Completed'
# (Reference 16)
prosp_sub[prosp_sub$loan_status %in% complete, 1] <- c('Completed')

# Now get rid of the factors that I don't use. 
# (Reference 15). Note that I ordered the levels from best case scenario
# (Completed) to worst case scenario (ChargedOff).  Loans that are 
# cancelled aren't really good or bad.

prosp_sub$loan_status <- factor(prosp_sub$loan_status, 
                    levels = c("Completed", "Current", "Past Due", 
                               "Defaulted", "Chargedoff", "Cancelled"))

# Get counts for each category

summary(prosp_sub$loan_status)

# Create the bar graph of proportion of borrowers (References 17 and 18)
ggplot(prosp_sub, aes(loan_status)) + 
  geom_bar(aes(y=..count../sum(..count..))) + ylab('Proportion of Borrowers')
```

From this plot, the category with the highest proportion of borrowers is clearly
the Current loans.  I lumped all the Past Due categories into one Past Due 
category since I only really care about Completed, Defaulted and Chargedoff
loans.  I also lumped the FinalPaymentInProgress category into the Completed
category since these loans are for all intents and purposes finished.  I did
some research and Chargedoff loans are Defaulted loans for which there is no
expectation of re-payment by the lenders (Reference 9).  Because of this, I am equally as interested in Chargedoff loans as I am those in Default. Completed 
loans clearly have a better proportion than the combination of Defaulted and Chargedoff loans (around 0.37 versus around 0.15).  It will be interesting to 
see how these proportions match up with those of employment status, credit score and the other variables.

```{r, echo=FALSE, message=FALSE, warning=FALSE, employment_status_graph}
# Plotting bar graph of proportions of borrowers with each employment status

# After graphing the initial version of the bar graph, I realize I need to 
# rotate the axis labels to be able to keep them from running into each other.
# (Reference 19).  I also graph the proportions as I did with loan status.


ggplot(prosp_sub, aes(employment_status)) + 
  geom_bar(aes(y=..count../sum(..count..))) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  ylab('Proportion of Borrowers')

# Get counts for each category
summary(prosp_sub$employment_status)

```

From the data available we can conclude that the information given isn't very helpful at possibly predicting whether a borrower will be successful in paying 
back their loans.  This is primarily because the categories are vague.  The 
highest proportion of borrowers fit into the "Employment" category.  This 
category is very vague.  It doesn't indicate whether the borrower is a Full-time 
or Part-time employee (there are actually categories for this).  It also doesn't indicate what their salary is.  There are also the categories of Other and a portion of the borrowers who did not list an employment status.  If I were to decide to examine this data further in the bivariate and multivariate portions 
of the analysis, I might get rid of the borrowers fitting in the Other and non-labeled categories.  I may construct a plot or two to see if there are relationships, but I think other variables I've chosen to focus on will give a better indication of success.  I picked this variable just to see if there may actually be an obvious pointer to a relationship but there isn't (for an obvious pointer to a relationship to exist, the proportions of one category would need 
to equal that of the Completed and Default/ChargedOff categories).  Even if the proportions were equal, the relationship may not be straight-forward between 
these two variables.

```{r, echo=FALSE,  message=FALSE, warning=FALSE, fig.height = 8, fig.width = 8, credit_score_range_lower}
# This chunk generates three versions of the histogram of lower range of credit scores.  I initially generate a basic histogram but then adjust it, zooming in 
# on the data and adjusting the binwidth.  I then graph all three of them on
# one grid.


# Plot the initial distribution of lower ranges of credit scores (Reference 20)


q1 <- ggplot(prosp_sub, aes(credit_score_range_lower)) + 
             geom_histogram() 

# Plot histogram while zooming in to the range of most of the data. I pick a max 
# of 880 because this value is the maximum given in the summary below (Reference 21). I used coord_cartesian so that I wouldn't shave off any data.

q2 <- ggplot(prosp_sub, aes(credit_score_range_lower)) + 
        geom_histogram() + coord_cartesian(xlim = c(375,880))

# Plot histogram while zooming in to the range of most of the data and picking a better binwidth (Reference 20)

q3 <- ggplot(prosp_sub, aes(credit_score_range_lower)) + 
       geom_histogram(binwidth = 1) + coord_cartesian(xlim = c(375,880))

# Read in grid extra to plot all these histograms on the same graph 
# (Reference 22)
library(gridExtra)

grid.arrange(q1, q2, q3, ncol = 1)

# Get summary of the low range of credit score

summary(prosp_sub$credit_score_range_lower)
```

From the three plots of the histograms, each more finely tuned than the next, 
I notice a few things.  The first is that there are some borrowers with a credit
score of zero.  I am going to see what the rest of the variables looks like
for some of these borrowers to see if I should discard them or if I should
keep them (if it was simply an error in entering the data).  The next is that
the data is pretty close to being normal.  From the summary given, the mean
is only 5.6 away from the median out of a range of 40 for the IQR.  Since
the mean is greater than the median however, it is skewed to the right.  I 
don't think it's enough to warrant performing a transformation on the data,
however.  The middle plot was bi-modal but changing the binwidth caused this
to disappear from the data. Another thing I noticed after changing the binwidth 
is that there appears to be only integer values of credit score in increments of 20. There do not appear to be values that aren't whole or that fall between 
values of 600, 620, 640 and so on (note that I confirmed this using the table command but didn't include it in the final code). In terms of lower range of 
credit score predicting the ability of a borrower to pay back their loans, there 
is a fairly steep drop off from around 640 to 620 in the number of borrowers.  
I would be interested to see what percentage of the borrowers to the left of the 
drop off are in each of the loan status categories.  I may look at how this 
variable changes with each loan status category using a faceted group of box 
plots in a bi-variate analysis.

```{r, echo=FALSE, message=FALSE, warning=FALSE, credit_score_range_lower0}

# See the first 6 rows of data for those with lower range of credit score of
# 0 (Reference 21)
head(subset(prosp_sub, credit_score_range_lower == 0))

```

Looking at the first 6 rows of data for those with a lower credit score range of 
0, I actually think the data is good and that is truly the lower range on the credit score.  The employment status and debt to income ratio is unknown for the first six rows of data, so these rows are probably cases where very little information is known about the borrowers so the risk for investors will be high.  The APR seems to be high for the first 6 rows of data so that would make sense 
for borrowers with bad credit scores.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 8, fig.width = 8, debt_to_income_ratio}
# This chunk generates three versions of the histogram of debt to income ratio.  
# I initially generate a basic histogram but then adjust it, zooming in 
# on the data and adjusting the binwidth.  I then graph all three of them on
# one grid.


# Plot the initial distribution of debt to income ratios (Reference 20)


q1 <- ggplot(prosp_sub, aes(debt_to_income_ratio)) + 
             geom_histogram() 

# Plot histogram while zooming in to the range of most of the data. I pick a max 
# of 1 since this seems to be where most of the data is (Reference 21). I used coord_cartesian so that I wouldn't shave off any data.

q2 <- ggplot(prosp_sub, aes(debt_to_income_ratio)) + 
        geom_histogram(binwidth = 0.1) + coord_cartesian(xlim = c(0,1))

# Plot histogram while zooming in to the range of most of the data and picking a better binwidth (Reference 20)

q3 <- ggplot(prosp_sub, aes(debt_to_income_ratio)) + 
       geom_histogram(binwidth = 0.005) + coord_cartesian(xlim = c(0,1))

# Plot all these histograms on the same graph (Reference 22)

grid.arrange(q1, q2, q3, ncol = 1)

# Get summary of the debt to income ratio

summary(prosp_sub$debt_to_income_ratio)
```

A few things stand out about these graphs.  The first is that there appears that there might be outliers close to a ratio of 10.  Again, I will check these out 
in the next R chunk to see if I should drop the data.  The second is that like 
the credit score graph, the data is skewed to the right.  This initially 
surprised me but it didn't after thinking on it some more.  I was initially thinking that the higher a borrower's debt to income ratio is, the less able 
they will be to pay back their loans.  But it actually may be that borrowers 
with higher incomes are able to take on much more debt and still be successful 
at paying back the loans (my guess is that there is a lot of overlap between the two graphs).  A significant portion of the debt for those with higher debt to income ratios is probably student loan debt.  It could very well be that 
borrowers with low debt to income ratios simply don't have the ability to take 
out loans because their income is low and because of this, they are the ones who will be less successful at paying back loans. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, debt_to_income_ratio5}

# See the first 6 rows of data for those with debt to income ratio greater than
# 5 (Reference 23)
head(subset(prosp_sub, debt_to_income_ratio >= 5))
```

After looking at the first few rows for borrowers with a debt to income ratio
greater than 5, I now think that this data should be left.  One of the reasons
borrowers may have an extremely high debt to income ratio is because they have
either lost their jobs but have fairly good credit scores or because they are
simply between jobs.  This would cause an extremely low income which would
explain the high debt to income ratios.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 8, fig.width = 8, borrower_apr}
# This chunk generates three versions of the histogram of borrower apr.  I initially generate a basic histogram but then adjust it, zooming in 
# on the data and adjusting the binwidth.  I then graph all three of them on
# one grid.


# Plot the initial distribution of borrower apr (Reference 20)


q1 <- ggplot(prosp_sub, aes(borrower_apr)) + 
             geom_histogram() 

# Plot histogram while zooming in to the range of most of the data. I pick a 
# max of 0.4 since this seems to be where most of the data is (Reference 21). I used coord_cartesian so that I wouldn't shave off any data.

q2 <- ggplot(prosp_sub, aes(borrower_apr)) + 
        geom_histogram() + coord_cartesian(xlim = c(0,0.4))

# Plot histogram while zooming in to the range of most of the data and picking a better binwidth (Reference 20)

q3 <- ggplot(prosp_sub, aes(borrower_apr)) + 
       geom_histogram(binwidth = 0.0005) + coord_cartesian(xlim = c(0,0.4))

# Plot all these histograms on the same graph (Reference 22)

grid.arrange(q1, q2, q3, ncol = 1)

# Get summary of the borrower_apr

summary(prosp_sub$borrower_apr)
```

The first thing to notice about this graph is that from the summary and the 
first two plots, the data appears to be skewed to the right.  This is confusing
to me because I was assumming that borrowers with good credit scores would
have lower APR's.  I was assuming that this graph would be skewed to the
left and be a reflection of the credit score and debt to income ratio
histograms.  A few things could be responsible for this result.  The first
is that there may not be a nice relationship between good credit score and 
borrower apr.  Perhaps a borrower has a good credit score but they have lost
their job which makes them a risky investment.  Or at the other end of the 
spectrum, perhaps a borrower has a bad credit history but has landed a good job such that prosper assigns a lower than expected interest rate.  And it is not 
even certain that there is a nice relationship between credit score and debt to income ratio as I guessed earlier.  I will have to graph these variables against each other to determine a discernible relationship.  Another observation 
pointing to the possibility that there may not be a nice relationship between borrower apr and credit score/debt to income ratio is that the mode on the 
borrower apr is significantly different from the mean and median.  From the 
third plot with a fine binwidth, the mode appears to be around 0.36 while the median and mean is around 0.21.  One final observation is that there are 
borrowers with apr's greater than 0.4.  I'm going to assume that these data 
points are accurate and are not outliers which have been incorrectly entered 
since I accepted the data I thought were outliers for debt to income ratio and credit score.  

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.height = 8, fig.width = 8, days_delinquent}
# This chunk generates three versions of the histogram of days delinquent.  I initially generate a basic histogram but then adjust it, zooming in 
# on the data and adjusting the binwidth.  I then graph all three of them on
# one grid.  Note that I only considered borrowers for which there was actually
# a days delinquent value (non-zero) and for which the loan had not been put in
# defaulted or charged off status.

# Create condition that the number of days delinquent must be greater than 0 
# and that the loan status must not be defaulted or charged off.
condition <- ((prosp_sub$days_delinquent > 0) & 
                (prosp_sub$loan_status != 'Defaulted') & 
                (prosp_sub$loan_status != 'Chargedoff'))

# Plot the initial distribution of days delinquent (Reference 20)

q1 <- ggplot(subset(prosp_sub, condition), aes(days_delinquent)) + 
             geom_histogram() 

# Plot histogram while zooming in to the range of most of the data. I pick a max 
# of 129 since this was the maximum number of days based on the summary below (Reference 21). I used coord_cartesian so that I wouldn't shave off any data.

q2 <- ggplot(subset(prosp_sub, condition), aes(days_delinquent)) + 
        geom_histogram() + coord_cartesian(xlim = c(0,129))

# Plot histogram while zooming in to the data and picking a better binwidth (Reference 20)

q3 <- ggplot(subset(prosp_sub, condition), aes(days_delinquent)) + 
       geom_histogram(binwidth = 1) + coord_cartesian(xlim = c(0,129))

# Plot all these histograms on the same graph (Reference 22)

grid.arrange(q1, q2, q3, ncol = 1)

# Get summary of the days_delinquent variable under the laid out conditions.

summary(subset(prosp_sub, condition)$days_delinquent)
```

For the days_delinquent variable, I only considered loans that had not been
charged off or defaulted since they are in a separate category and will have
large values.  I also only considered cases for which there was a non-zero value
for days_delinquent.  Had I considered all the borrowers, a disproportionally 
large number of them would have been zero days' delinquent and I wouldn't be 
able to gain any insights on borrowers who hadn't yet defaulted or been charged
off on their loans who also are delinquent in paying their loans.  From these
distributions, the data is definitely skewed to the right with the median at 27 
days.  Most of the cases are below 50 days.  It's hard to know which variables
may predict the number of days' delinquent since thus far I have considered all
the data and not subsetted the data as I did in this R chunk.  It may also be
good in the future to consider the ratio of days' delinquent to number of days 
since the loan originated.  This might be a better indicator of the relative
success of borrowers paying back their loans.  A borrower may be 30 days'
delinquent but only 2 months into paying back the loans while another borrower 
may be 30 days' delinquent but be halfway through paying back the loans (a 
matter of years).  The delinquency for the second borrower isn't going to be 
viewed in the same way as that of the first.  This variable is a secondary independent variable to that of loan status so I'm not as concerned at gaining insights into what may predict these numbers.

```{r, echo=FALSE, message=FALSE, warning=FALSE, is_homeowner}
# Plotting bar graph of proportions of borrowers that are and are not homeowners
# (References 17 and 18)

ggplot(prosp_sub, aes(is_homeowner)) + 
  geom_bar(aes(y=..count../sum(..count..))) + 
  ylab('Proportion of Borrowers')

# Get counts for each category

summary(prosp_sub$is_homeowner)
```

From the bar chart above showing proportions of borrowers who own a home, we 
can see that there is minimal difference between the conditions.  The proportion
of borrowers who own a home is slightly greater than 50%.  It's hard to
determine how well this variable predicts loan payback success rate without
using bi-variate and multi-variate graphic capabilities in R.  I will facet loan 
status over each of these conditions to see which condition has a higher level 
of success and failure.  It would make sense that borrowers owning a home will 
be more likely to payback their loans than those that do not own a home. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, subset_date}
# This R chunk subsets the prosper data such that only data after July 2009
# is considered.  This data will actually have a prosper score and rating.

# Several of the loans had NA for prosper scores.  These loans originated
# in or prior to July 2009.  By subsetting the data to get rid of rows for 
# which the prosper score is NA, I am effectively only considering loans 
# which originated after July 2009 (References 13 and 24).

prosp_2009 <- subset(prosp_sub, !is.na(prosper_score))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, prosp_2009_combine}
# This R chunk combines some of the prosper scores into one category so that
# I am not considering 11 categories which seem to be a bit too much.

# Create a variable which contains all the possible prosper score low
# categories.  They will be lumped into one category called '1-3'  
# (Reference 12). Also create a variable which contains all the possible 
# prosper score high categories, they will be lumped into one category called
# '9-11'.

prosper_score_low <- c('1', '2', '3')
prosper_score_high <- c('9', '10', '11')

# First I need to turn the variable prosper_score into a factor variable. I 
# will do this in both the prosp_sub and prosp_2009 dataframes (Reference 15).

prosp_sub$prosper_score <- factor(prosp_sub$prosper_score)
prosp_2009$prosper_score <- factor(prosp_2009$prosper_score)

# Now I need to add two factor levels to the factor variable prosper_score (Reference 15). I will do this for both prosp_2009 and prosp_sub.

prosp_sub$prosper_score <- factor(prosp_sub$prosper_score, 
                    levels = c(levels(prosp_sub$prosper_score), "1-3", "9-11"))

prosp_2009$prosper_score <- factor(prosp_2009$prosper_score, 
                    levels = c(levels(prosp_2009$prosper_score), "1-3", "9-11"))

# Now for all categories contained in prosper_score_low and prosper_score_high
# set them to '1-3' and '9-11', respectively (Reference 16).

prosp_sub[prosp_sub$prosper_score %in% prosper_score_low, 6] <- c("1-3")

prosp_2009[prosp_2009$prosper_score %in% prosper_score_low, 6] <- c("1-3")

prosp_sub[prosp_sub$prosper_score %in% prosper_score_high, 6] <- c("9-11")

prosp_2009[prosp_2009$prosper_score %in% prosper_score_high, 6] <- c("9-11")

# Now get rid of the factors that I don't use. 
# (Reference 15). Note that I ordered the levels from highest risk (1-3) to
# lowest risk (9-11).

prosp_sub$prosper_score <- factor(prosp_sub$prosper_score, 
                    levels = c("1-3", "4", "5", "6", "7", "8", "9-11")) 

prosp_2009$prosper_score <- factor(prosp_2009$prosper_score, 
                    levels = c("1-3", "4", "5", "6", "7", "8", "9-11")) 
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, prosper_score_graph}
# This R chunk creates a bar graph of proportions of each borrowers
# with the prosper score levels I created in the previous R chunk

# Create the bar graph of proportion of borrowers (References 17 and 18)
ggplot(prosp_2009, aes(prosper_score)) + 
  geom_bar(aes(y=..count../sum(..count..))) + ylab('Proportion of Borrowers')

# Get counts for each category

summary(prosp_2009$prosper_score)
```

Based on the bar chart shown, the combinations of prosper scores 1-3 have the
largest proportion of borrowers while the combinations of prosper scores 9-11
have the second largest proportion of borrowers.  I decided to combine the 
totals for 1-3 and 9-11 into one category so that I wouldn't be considering 11
different categories which seemed to be too much.  It should be noted that
scores of 4 and 6 would have had the highest proportions if I hadn't done this.
This would be reasonable since most users will have around an average risk
associated with the investment (prosper scores of 1 and 11 would have had the lowest and second lowest proportions, respectively).  In terms of prosper scores predicting whether a loan will be defaulted, chargedoff or completed, it is hard 
to tell.  It is worth mentioning that around the same proportion of borrowers 
assigned a rating between 9-11 also either defaulted or were chargedoff with 
their loans based on the loan status graph previously, but that data considered
all loans and not just those after July 2009 (the bar chart above is only for
data post July-2009).  I will need to use R's bi-variate graphing capabilities 
to see if a relationship exists, I suspect it will. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, prosper_rating_graph}
# This R chunk creates a bar graph of proportions of each borrowers
# with the possible prosper rating levels

# Change order of levels such that they go from highest risk (worst, HR) to 
# lowest risk (best, AA) (Reference 15)
prosp_sub$prosper_rating <- factor(prosp_sub$prosper_rating, 
                    levels = c("HR", "E", "D", "C", "B", "A", "AA")) 

prosp_2009$prosper_rating <- factor(prosp_2009$prosper_rating, 
                    levels = c("HR", "E", "D", "C", "B", "A", "AA")) 

# Create the bar graph of proportion of borrowers (References 17 and 18)
ggplot(prosp_2009, aes(prosper_rating)) + 
  geom_bar(aes(y=..count../sum(..count..))) + ylab('Proportion of Borrowers')

# Get counts for each category

summary(prosp_2009$prosper_rating)
```

From the above bar chart, it appears that the prosper rating with the highest
proportion of borrowers is that with a rating of C.  As with the prosper scores,
this is reasonable since this is the average prosper rating.  Most borrowers
will not have excellent or bad credit but have average credit.  The two ratings
with the lowest and second lowest proportion of borrowers are the AA and HR
ratings, respectively.  As I mentioned with the prosper score, I will need to
use R's bi-variate graphing capabilities to determine the relationship between
prosper score and loan status. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, loan_status_comparison}
# Plotting two bar graphs of loan_status; one for the best possible combination
# of prosper rating and prosper score (AA, 9-11).  The other for the worst
# possible combination (HR, 1-3).  This is to see if there may be a relationship
# between these three variables.

# Set conditions
best_combo <- (prosp_2009$prosper_score == "9-11" & 
                 prosp_2009$prosper_rating == "AA")
worst_combo <- (prosp_2009$prosper_score == "1-3" & 
                  prosp_2009$prosper_rating == "HR")

# Create the bar graph of proportion of borrowers for best possible combination (References 17 and 18)

q1 <- ggplot(subset(prosp_2009, best_combo), aes(loan_status)) + 
  geom_bar(aes(y=..count../sum(..count..))) + ylab('Proportion of Borrowers')

q2 <- ggplot(subset(prosp_2009, worst_combo), aes(loan_status)) + 
  geom_bar(aes(y=..count../sum(..count..))) + ylab('Proportion of Borrowers')

grid.arrange(q1, q2, ncol = 1)
```

The first plot in the visual directly above is the loan status for the best 
possible combination of prosper rating and score (AA and 9-11) while the second plot is the loan status for the worst possible combination of prosper rating and score (HR and 1-3). Clearly there is a higher precedence of loans being past 
due, defaulted and charged off for the worst combination than the best 
combination.  There is also a higher percentage of loans which are completed in 
the best combination versus the worst combination.  This would seem to point to 
a relationship between these three variables.  I will be examining this further 
in the bi-variate and multi-variate portion of my analysis.  

```{r, echo=FALSE, message=FALSE, warning=FALSE,  days_delinquent_comparison}
# Plotting two histograms of days_delinquent; one for the best possible 
# combination of prosper rating and prosper score (AA, 9-11).  The other for the 
# worst possible combination (HR, 1-3).  This is to see if there may be a 
# relationship between these three variables.

# Set conditions. Again, I limit things to cases where the days_delinquent
# is greater than 0 and the loan status is not defaulted or charged off.
best_combo <- (prosp_2009$prosper_score == "9-11" & 
                 prosp_2009$prosper_rating == "AA" & 
                 prosp_2009$days_delinquent > 0 &
                 prosp_2009$loan_status != "Defaulted" &
                 prosp_2009$loan_status != "Chargedoff")
worst_combo <- (prosp_2009$prosper_score == "1-3" & 
                 prosp_2009$prosper_rating == "HR"& 
                 prosp_2009$days_delinquent > 0 &
                 prosp_2009$loan_status != "Defaulted" &
                 prosp_2009$loan_status != "Chargedoff")

# Create the histogram of proportion of borrowers for best and worst possible
# combinations.  I pick a max of 125 since all the data is below this number (References 20 and 21).  I used coord_cartesian so that I wouldn't shave off any data.

q1 <- ggplot(subset(prosp_2009, best_combo), aes(days_delinquent)) + 
        geom_histogram(binwidth = 1) + coord_cartesian(xlim = c(0,125))

q2 <- ggplot(subset(prosp_2009, worst_combo), aes(days_delinquent)) + 
        geom_histogram(binwidth = 1) + coord_cartesian(xlim = c(0,125))

# Plot the charts on a grid (Reference 22)

grid.arrange(q1, q2, ncol = 1)
summary(subset(prosp_2009, best_combo)$days_delinquent)
summary(subset(prosp_2009, worst_combo)$days_delinquent)
```

The above two histograms are the distributions of days delinquent under the best (AA and 9-11) and worst (HR and 1-3) possible combination of prosper scores and ratings as I did with the loan status variable.  As I did in the original plots 
of days delinquent, I limited the data to cases where the days delinquent was greater than 0 and to cases where the loan had not been charged off or 
defaulted.  Clearly there are fewer cases of days delinquent for the best combination than the worst combination.  Looking at the summary however, the 
average number of days delinquent is greater for the best case scenario than 
the worst case scenario.  A couple of things should be kept in mind.  The first 
is that it may be better to create a variable where the number of days 
delinquent is divided by the number of days since the origination of the loan.  
As I mentioned before, someone may be thirty days delinquent on a loan but be really close to finishing paying off the loan.  The second is that the number of borrowers delinquent being greater in the worst combination case may simply be because Prosper assigned more borrowers this combination than they did borrowers the best combination.  It could be that prospers ratings and scores are good at predicting loan status but not good at predicting days delinquent.  Again, I'm 
not as concerned about predicting this variable but it is something to look at 
in future analyses. 

```{r echo=FALSE, message=FALSE, warning=FALSE, resources}
# Resources - code

# Reference 1: http://ggplot2.org/

# Reference 2: https://yihui.name/knitr/

# Reference 3: https://dplyr.tidyverse.org/

# Reference 4: https://stackoverflow.com/questions/29189663/trying-to-
# publish-an-r-notebook-and-keep-getting-the-same-error-error-in-contri

# Reference 5: http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html

# Reference 6: https://stackoverflow.com/questions/5595512/what-is-the-
# difference-between-require-and-library

# Reference 7: http://rfunction.com/archives/1001

# Reference 8: https://stackoverflow.com/questions/6329962/split-code-over-
# multiple-lines-in-an-r-script

# Reference 9: http://rprogramming.net/read-csv-in-r/

# Reference 10: https://stat.ethz.ch/R-manual/R-devel/library/base/html/dim.html

# Reference 11: https://www.rdocumentation.org/packages/utils/versions/3.4.3/
# topics/str

# Reference 12: https://stackoverflow.com/questions/11488820/why-use-c-to-
# define-vector

# Reference 13: https://www.rdocumentation.org/packages/base/versions/3.5.0/
# topics/subset

# Reference 14: http://www.sthda.com/english/wiki/renaming-data-frame-columns-
# in-r

# Reference 15: https://stats.idre.ucla.edu/r/modules/factor-variables/

# Reference 16: https://stackoverflow.com/questions/32845303/using-logical-
# operators-with-characters

# Reference 17: https://stackoverflow.com/questions/36604127/creating-a-bar-
# plot-with-proportions-on-ggplot

# Reference 18: http://ggplot2.tidyverse.org/reference/geom_bar.html

# Reference 19: http://tallman-world.tumblr.com/post/89964965867/rhow-to-
# rotate-axis-labels-in-ggplot2

# Reference 20: http://ggplot2.tidyverse.org/reference/geom_histogram.html

# Reference 21: http://ggplot2.tidyverse.org/reference/coord_cartesian.html

# Reference 22: https://www.rdocumentation.org/packages/gridExtra/versions
# /2.3/topics/arrangeGrob

# Reference 23: https://stackoverflow.com/questions/2667673/select-first-4
# -rows-of-a-data-frame-in-r

# Reference 24: https://www.rdocumentation.org/packages/base/versions
# /3.5.0/topics/NA


# NOTE: This is the Resources list only for the code chunks.  I am placing this list inside of this chunk so that they do not appear when knitting the HTML 
# file (for users who want to do that at the end). The only resources I want to appear when knitting the HTML file are those referenced inside the file and 
# meant to be visible (outside the R chunks).
```

# Resources

Reference 1: https://www.prosper.com/home/about

Reference 2: https://www.kaggle.com/jschnessl/prosperloans/feed

Reference 3: https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0

Reference 4: https://www.investopedia.com/articles/00/091800.asp

Reference 5: https://www.prosper.com/plp/general-prosper_score/

Reference 6: https://www.investopedia.com/terms/d/delinquent.asp

Reference 7: https://www.prosper.com/Downloads/Legal/Prosper_Prospectus_2018-04-30.pdf

Reference 8: https://www.investopedia.com/terms/a/apr.asp

Reference 9: https://help.lendingclub.com/hc/en-us/articles/216127747-What-is-the-difference-between-a-loan-that-is-in-default-and-a-loan-that-has-been-charged-off-